---
title: "PrinStat_RCode"
author: "Nathan Laroy"
date: "2023-09-29"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r init}
library(tidyverse)
library(car)
```

## 1) Read Data

```{r}
#classic way
dat <- read.csv(file.choose(), header = TRUE, sep = ";")

#tidyverse way
dat <- read_csv2(file.choose(), col_names = TRUE, na = character())
```

## 2) Data meta-descriptives (structure, dimensions, names, head, tail ...)

```{r}
#head (top rows)
head(dat)

#tail
tail(dat)

#names
names(dat)

#dimensions
dim(dat)

#structure --> characterizes var type and contents of each element of "names", yields also dim
str(dat)
```

To look at specific columns (i.e. vars) of dataset:

```{r}
#classic way
dat$Gender

#alternative ==> compatible with *tidyverse* %>% pipe operator 
with(dat, Gender)

#example with %>%
dat %>% pull(Gender)
```

To subset from dataset into dataframe-type:

```{r}
#classic way (does not retain column name, becomes "dat.Gender" instead of "Gender")
subsetted <- data.frame(dat$Gender)

#tidyverse way (retains column name)
dat %>% select(Gender)
```

*Turn all factors into factor objects, such that R treats them as such!* Use "within" to alter element of data

```{r}
#example
dat <- within(dat, Gender <- factor(Gender))

#is equivalent to
dat$Gender <- as.factor(dat$Gender)
```

## UNIVARIATE DESCRIPTIVES

Remember: it is good practice to use with(<dataframe object>, <var of interest>) instead of ...\$..., because it is compatible with tidyverse pipeline operators

### Categorical Variables

```{r}
# General structure of dataset (here: subset)
with(dat, str(Gender))

# table with absolute counts
with(dat, table(Gender)) #prints to console
#or
dat %>% count(Gender) #does not print to console

# table with proportions
with(dat, prop.table(table(Gender)))

# to table without null or NA, use "exclude = ..." 
with(dat, table(Gender, exclude="")) #exclude all instances of ""

# to explicitly *include* NA in table/prop.table :: "exclude = NULL"
with(dat, prop.table(table(Gender, exclude=NULL)))
```

Change all "" into NA:

```{r}
dat <- within(dat, Gender <- factor(na_if(Gender, "")))
```

Visualize if any observations are "NA" using Booleans:

```{r}
with(dat, is.na(Gender))
```

Visualize cat.var using **bar charts**

```{r}
with(dat, plot(Gender, xlab="Gender",ylab="counts",main="bar chart"))

# to explicitly include NA, specify names.arg
tab <- with(dat, table(Gender, exclude=NULL))
barplot(tab, ylab = "Absolute Frequency",
        names.arg = c("female", "male", "NA"), xlab = "Gender")

# prop table with NA included: (note that proportions must be calculated manually, otherwise raises error)
barplot(tab / length(dat$Gender), ylab = "Percentage",
        names.arg = c("female", "male", "NA"), xlab = "Gender")

# !! if data contains missing values (NA), then plot() will discard them :: use barplot() instead !!
```

Of course, bar charts can also be made with tidyverse's subpackage **ggplot2**. A difficulty is that its plots cannot be included in a pipe chain (with `%>%`). Instead, they use their own `+` operator to add various "layers" to a plot:

```{r}
plt_Gender <- ggplot(data = dat, mapping = aes(x = Gender)) +
  geom_bar(fill = "skyblue")
plt_Gender
```

A(nother) quirk of **ggplot2** is that there is no straightforward way to create bar charts with relative frequencies:

```{r}
ggplot(data = dat, mapping = aes(x = Gender)) +
  geom_bar(fill = "skyblue", mapping = aes(y = (..count..) / sum(..count..))) +
  labs(y = "percent")
```

#### change display parameters globally for visualizing

R does not always display all level names (or level data) if there are too many. A solution is a horizontal bar chart but then we have to ensure that the *figure margin* is wide enough so the levels are readable. We can change the figure margins with the function `par()`:

```{r}
# use 'par' to set parameters

# here: mar = c(b, l, t, r) == bottom, left, top, right margin
# numbers represent number of lines left of graph edge 
# (i.e., here, for hor bars, leave 12 lines width free for bar labels)
par(mar = c(5, 12, 4, 2) + 0.1)
barplot(tab, horiz = TRUE, las = 1, cex.names = 0.6)
#las -> orientation of axis labels (0 = always parallel to axis, 1 = always horizontal, 2 = always perpendicular to axis, 3 = always vertical)


# ALWAYS RESET TO DEFAULT MARGINS !!
par(mar = c(5, 12, 4, 2) + 0.1)
```

### Numeric/Continuous variables

In case the variables `Distance.to.class` and/or `Time.to.get.here` contain non-numeric characters such as e.g. `km` or `mins` (respectively), they will be read in as character columns. In that case you need to first convert both variables as follows:

```{r}
dat <- within (dat, {
  Distance.to.class <- as.numeric(gsub("[[:alpha:]]", "", Distance.to.class))
  Time.to.get.here <- as.numeric(gsub("[[:alpha:]]", "", Time.to.get.here))
})
# "[[:alpha:]]" means "any of the alphabetic characters"
# so, any of alphabet will be substituted by "", which is empty string
```

How many missing values?

```{r}
with(dat, sum(is.na(Distance.to.class)))
```

Rearrange data according to observation values:

```{r}
#ascending arrangement
dat %>% arrange(Distance.to.class)

#descending arrangement
dat %>% arrange(desc(Distance.to.class))
```

R of course has many functions for computing **numerical summaries** (i.e. descriptive statistics). By default, however, they return `NA` if there are any `NA`'s in your data. If you want to have the numerical summary for the non-missing data, then each of these functions has to be used with the argument `na.rm = TRUE`:

```{r}
#example
with(dat, min(Distance.to.class))
with(dat, min(Distance.to.class, na.rm = T))
```

Various summary statistics can be combined in a vector:

```{r}
with(dat, 
     c(mean = mean(Distance.to.class, na.rm = TRUE),
       var = var(Distance.to.class, na.rm = TRUE),
       sd = sd(Distance.to.class, na.rm = TRUE))
     )
```

The tidyverse counterpart of this is to work with the `summarise()` function (of the **dplyr** subpackage):

```{r}
dat %>% summarise(
  mean = mean(Distance.to.class, na.rm = TRUE),
  var = var(Distance.to.class, na.rm = TRUE),
  sd = sd(Distance.to.class, na.rm = TRUE))
```

Quantiles example:

```{r}
with(dat, quantile(Distance.to.class, probs = (1:10) / 10, na.rm = TRUE))
# "/ 10" to get decimals (otherwise not "proportions")

# boxplot quantiles + mean + range
with(dat, summary(Distance.to.class, na.rm = TRUE))
```

**Graphical summaries** are provided by the histogram, boxplot or QQ-plot.

**Histogram**:

```{r}
# put "freq = F" to get absolute counts on y-axis
with(dat, hist(Distance.to.class, freq = T))

# to include density line
# to specify breaks (you can use numerical, or specify based on levels in factor by referring to factor)
with(dat, {
  hist(Distance.to.class[-1], freq = FALSE, breaks = 20)
  lines(density(Distance.to.class[-1], na.rm = TRUE))
        })
```

Get indices of specified values using `which()` (or `match()` for specific equalities). Subset using indices by `...[-...]`.

```{r}
with(dat, which(Distance.to.class > 7000))

# subset example in hist()
with(dat, hist(Distance.to.class[-1], freq = T))
```

The tidyverse counterpart to excluding observations is to use the function `filter()` (from the subpackage **dplyr**). Since we cannot include base R functions (like `hist()`) in pipe chains (using `%>%`), we have to create a new data set:

```{r}
sub <- dat %>% filter(Distance.to.class < 10000)
with(sub, hist(Distance.to.class, freq = FALSE))
```

**Boxplot**:

```{r}
with(dat, boxplot(Distance.to.class, ylab = "distance to class", 
                  main = "boxplot of distance"))

# horizontal and exclude outliers:
with(sub, boxplot(Distance.to.class, horizontal = TRUE, outline = FALSE))
# "outline = F" will remove outliers from plot
```

The function `Boxplot()` from the package **car** is more informative than the base R function in that it plots the row ID's of the outliers (execute `install.packages("car")` if you have not yet installed the package):

```{r}
with(dat, Boxplot(Distance.to.class, ylab = "distance to class", 
                  main = "boxplot of distance"))
```

**QQ-plot**: = Check for normality (relevant for various statistical tests, e.g. t-test).

With outliers:

```{r}
with(dat, {
  qqnorm(Distance.to.class)
  qqline(Distance.to.class)
})
```

**multiple charts at once**

```{r}
# 1) set parameter space: 1 row, 2 col
par(mfrow = c(1, 2))  # split the screen into a matrix with 1 row and 2 columns

# 2) display boxplot
with(sub, {
  hist(Distance.to.class, xlab = "Distance to class", main = "")
  boxplot(Distance.to.class)
})

# ALWAYS RESET PARAMETERS TO DEFAULT !!
par(mfrow = c(1, 1))
```

What about even more plots together?

```{r}
# set layout for next N figures (every N specifies indexed figure, so "1" is first picture)
# in matrix(c(1,1,2,3)) means "1" figure occupies top two cells, 
# and "2" and "3" occupy third and fourth quadrants
layout(matrix(c(1, 1, 2, 3), nrow = 2, ncol = 2, byrow = TRUE))
with(sub, {
  qqnorm(Distance.to.class)
  qqline(Distance.to.class)
  hist(Distance.to.class, xlab = "Distance to class", main = "")
  boxplot(Distance.to.class)
})
```

Time variables are numeric variables with special attributes in R:

```{r}
with(dat, str(Date.of.Birth))
```

```{r}
with(dat, Date.of.Birth)
```

They can be converted to date objects, i.e. objects with class `Date`:

```{r}
?as.Date    # You may have to change the format argument:
```

You may have to change the `format` string in the following code (e.g. `"%m/%d/%Y"` instead of `"%Y-%m-%d"`):

```{r}
dat <- within(dat, Birth.date <- as.Date(dat$Date.of.Birth,
                                 format = "%m/%d/%Y")
             )  # Change to format = "%m/%d/%Y" if the following produces NA's:
with(dat, Birth.date)

# histogram
with(dat, hist(Birth.date, breaks = "years"))

# specify format
with(dat, format(Birth.date, "%Y"))
```

## MULTIVARIATE DESCRIPTIVES

### Categorical x Categorical

```{r}
tab <- with(dat, table(Gender, Nationality, exclude = NULL))
prop.table(tab)

# change y and x axis by transposing
t(tab)
```

To display bar chart as conditional distributions, use `beside = T`.

```{r}
barplot(tab, beside = TRUE)    # use beside argument!
```

In **ggplot2** you put a variable on the x-axis by means of the aesthetics, but you first have convert the frequency table to a data frame (or tibble). The count variable in this data frame needs to be incorporated in the (mapping to) aesthetics but this also means that you have to use the `identity` stat (instead of the default `count` stat). Bar charts in **ggplot2** are also stacked by default so you plot the bars side-by-side with the argument `position = "dodge"`:

```{r}
df2 <- as.data.frame(tab)
df2
```

```{r}
ggplot(data = df2, mapping = aes(x = Gender, fill = Belgian, y = Freq)) +
  geom_bar(stat = "identity", position = "dodge") 
```

By the way, the data frame `df2` can always be (re)converted to a table by means of the function `pivot_wider()` of (the subpackage **tidyr** of) tidyverse. That function has many arguments but the most important two are:

-   `names_from`: The character column which should become the new columns
-   `values_from`: The numeric column which contains the values for the cells

```{r}
df3 <- df2 %>%
  pivot_wider(names_from = "Belgian", values_from = "Freq")
df3
```

The function `pivot_longer()` performs the opposite operation: it rearranges various columns underneath each other. You should should specify which columns should be rearranged and the arguments `names_to` and `values_to` have the opposite meanings to `names_from` and `values_from` in `pivot_wider()`:

```{r}
df4 <- df3 %>%
  pivot_longer(!Gender, names_to = "Belgian", values_to = "Freq")
df4
```

### Categorical x Continuous

```{r}
with(dat, aggregate(Time.to.get.here ~ Gender, FUN = mean))
```

Base R's `aggregate()` function has a counterpart in combining the `group_by()` function with `summarise()` of (the subpackage **dplyr** of) tidyverse:

```{r}
dat %>%
  group_by(Gender) %>%
  summarise(avg = mean(Time.to.get.here, na.rm = TRUE),
            med = median(Time.to.get.here, na.rm = TRUE),
            std = sd(Time.to.get.here, na.rm = TRUE))
```

Special case: the **evolution of a variable** (over time)

```{r}
with(sub, plot(Birth.date, Gender))
```

Add a smoother to gain more insight in `Gender` gap conditional on `Age` category (using **LOWESS** or **LOESS** = LOcally WEighted Scatterplot Smoothing):

```{r}
plot(as.numeric(Gender) ~ Birth.date, data = sub, yaxt = "n",
     xlab = "Birth date", ylab = "Gender")
with(sub, lines(loess.smooth(Birth.date, as.numeric(Gender)),
                        col = "red"))
abline(h = 1.5, lty = 2)
axis(2, at = c(1, 2), labels = levels(sub$Gender), las = 1)
```

### Continuous x Continuous

```{r}
with(sub, {
  plot(Birth.date, Age)    # default: scatterplot
  lines(loess.smooth(Birth.date, Age), 
        col = "red", lwd = 2)
})
```

Draw smoothed line 'fitted' to data:

```{r}
plot(Time.to.get.here ~ Distance.to.class, data = sub)
with(sub, lines(loess.smooth(Distance.to.class, Time.to.get.here),
                col = "red", lwd = 2))
```

Ditto, but for factor levels seperately:

```{r}
# plotting object
# col = *levels of factor*
plot(Time.to.get.here ~ Distance.to.class, data = sub, 
     col = sub$Means.of.transportation, pch = 19)

# add legend
legend(x = "topleft", col = 1:3,
       legend = levels(sub$Means.of.transportation), pch = 19)

# add smoothed line for each level of factor of interest
with(subset(sub, Means.of.transportation == "bike"),
     lines(loess.smooth(Distance.to.class, Time.to.get.here),
           col = 1, lwd = 2))
with(subset(sub, Means.of.transportation == "car / motorbike"),
     lines(loess.smooth(Distance.to.class, Time.to.get.here),
           col = 2, lwd = 2))
with(subset(sub, Means.of.transportation == "public transport"),
     lines(loess.smooth(Distance.to.class, Time.to.get.here),
           col = 3, lwd = 2))
```

### Continuous x Continuous x Continuous

As an example, let's add Age instead of means of transportation as the third variable. Instead of color coding, we can vary the size of the circles according to `Age` (larger circles = older students). This is called a **bubble plot**:

```{r}
plot(Time.to.get.here ~ Distance.to.class, data = sub, pch = "")
with(subset(sub, !is.na(Age)),
     symbols(Distance.to.class, Time.to.get.here, circles = Age,
             inches = 0.1, add = TRUE))
```
